---
title: "Tycho Data Level1 - National Incidence Rate"
format:
  html:
    fig-width: 34
    fig-height: 30
    toc: true
    page-layout: full
  pdf:
    fig-width: 24
    fig-height: 12
    toc: true
  docx:
    fig-width: 24
    fig-height: 30
    toc: true
---

```{r, system, echo=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
source("./utils.R")
```

## Data

Based on Tycho Data Version 1 from US MMWR and its predecessors that was 
digitized prior to the 
[NEJM article, Van Panhuis et al. 2013](https://doi.org/10.1056/NEJMms1215400). 
"Level 1" data: data used in the original NEJM publication mentioned above. 
Some processing and incidence rates were calculated, and some missing data were 
imputed (more information in the 
[associated supplement](https://doi.org/10.1056/NEJMms1215400))

Data are directly downloadable from 
[Project Tycho Website](https://www.tycho.pitt.edu/data/) or 
[Zenodo](https://zenodo.org/records/12608992)


### Census Data
Census data From:

- [Data from 1910 to 1990](https://www.census.gov/data/tables/time-series/demo/popest/1970s-state.html) 
- [Data from 1990 to 2000](https://www.census.gov/data/tables/time-series/demo/popest/1990s-state.html) 
- [Data from 2000 to 2010](https://www.census.gov/data/tables/time-series/demo/popest/intercensal-2000-2010-state.html)
- [Data from 2010 to 2020](https://www.census.gov/data/tables/time-series/demo/popest/intercensal-2010-2020-state.html) 

### Vaccination

Extracted from the 
[NEJM article, Van Panhuis et al. 2013](https://doi.org/10.1056/NEJMms1215400):

- Hepatitis A: 1995
- Measles: 1963
- Mumps: 1968 (1967, but for plotting reason change to 1968 (first date))
- Pertussis: 1948
- Polio: 1955
- Rubella: 1969
- Smallpox: NA, (1800s)
- Diphtheria: 1923

## Process

Calculated by adding all the state or city (for Diphtheria) together (excluding
US territories) per epi week and disease, and calculating the incidence rate 
per 100,000 people by calculating: 

$$\frac{nbr\ of\ cases}{population\ size} \times 100,000$$

The original Tycho files contained two "outliers" peaks: one for Hepatitis A 
on the week 52 of 2010 and one from Smallpox on the week 36 of 1948.

- For Hepatitis A, it seems to be a “print” error in the CDC source data. 
  It has been corrected but the correction has not been integrated to Tycho yet.
  Please see the erratum in the 
  [source data](https://www.cdc.gov/mmwr/PDF/wk/mm5951.pdf), page 1710.
  The corrected table is also availalbe an the 
  [CDC.gov](https://www.cdc.gov/mmwr/preview/mmwrhtml/mm6008a9.htm?s_cid=mm6008a9_w)
  website.
  To fix that issue, we removed the associated week for Hepatitis A for all 
  Tycho data

- For Smallpox, it seems to be “data entry” error, correct table is 
  available [here](https://pmc.ncbi.nlm.nih.gov/articles/PMC1995387/). 
  As the tables contains multiple disease information, the week 36 of 1948 has 
  been removed for all disease on all Tycho data.


## Visualization

*Notes*: 
  
- For Polio:
  "In the United States, the last cases of indigenously transmitted wild 
  poliovirus disease were reported in 1979, and the last imported case of
  poliomyelitis caused by wild type poliovirus was reported in 1993" (from
  [Chapter 12: Poliomyelitis, Manual for the Surveillance of Vaccine-Preventable Diseases. Adriana S. Lopez et ak, 2024](https://www.cdc.gov/surv-manual/php/table-of-contents/chapter-12-poliomyelitis.html?CDC_AAref_Val=https://www.cdc.gov/vaccines/pubs/surv-manual/chpt12-polio.html))
  
- For Smallpox:
  "By the time the Intensified Eradication Program began in 1967, smallpox was 
  already eliminated in North America (1952) and Europe (1953). Cases were still
  occurring in South America, Asia, and Africa (smallpox was never widespread in
  Australia). The Program made steady progress toward ridding the world of this 
  disease, and by 1971 smallpox was eradicated from South America, followed by
  Asia (1975), and finally Africa (1977)." (from 
  [CDC](https://www.cdc.gov/smallpox/about/history.html))
  
- For Pertussis: "Weekly reports for pertussis were unavailable from 1955 to 
  1974"[NEJM article, Van Panhuis et al. 2013](https://doi.org/10.1056/NEJMms1215400)

- New York and New York City: "After 1963, New York City and Upstate New York 
  were split into separate jurisdictions by the notifiable disease surveillance
  system" 
  [NEJM article, Van Panhuis et al. 2013](https://doi.org/10.1056/NEJMms1215400). 
  Since after 1963, both Upstate New York and New York City were reported as 
  if they were states, they were originally treated as states for data 
  reported via “state” reporting systems in version 1 of Project Tycho.

### National Incidence

(legend inspired from NEJM paper)

```{r, echo=FALSE}
#| fig-cap: "Weekly incidence rates per 100,000 population are shown for the entire country. The year of vaccine licensure is indicated by a vertical red line. The smallpox vaccine was introduced in the United States around 1800, so its licensure is not indicated."
#| layout-nrow: 2
df <- read.csv("../process_data/national_incidence_rate_lvl1.csv") %>%
  dplyr::mutate(date = as.Date(date), vaccine = as.Date(vaccine))
## Plot
disease_sel <- unique(df$disease)
## Plot
plot_national_incidence(dplyr::filter(df, disease %in% disease_sel[1:4]),
                        save = FALSE, font_size = 22, font_face = "bold",
                        line_width = 2, dpi = 320)
plot_national_incidence(dplyr::filter(df, disease %in% disease_sel[5:8]),
                        save = FALSE, font_size = 22, font_face = "bold",
                        line_width = 2, dpi = 320)
```

